Local PCA Shows How the Effect of Population Structure Differs Along the Genome

### 2017-06-21 writeup/review-responses.tex

\pagebreak

%%%%%%%%%%%%%%
\reviewersection{AE}

\begin{quote}
    Please understand that incremental changes will not be sufficient.  Adding
    simulations to strengthen key claims will be necessary, particularly addressing
    the impacts of mutation rate and recombination rate variation with more depth,
    the concern regarding PC switching (Reviewer 1), and the concern regarding the
    impacts of variation in missingness by sub-population (Reviewer 2).
\end{quote}

Thanks for the positive feedback and the useful suggestions.
We agree that more extensive exploration 
using simulations would help bolster understanding of the method,
and have now done so.
This took a substantial amount of work, because genome-scale forwards-time simulations
with a large number of loci under selection
is at or beyond the current limits of computation,
depending on the number of individuals simulated.

@@

    The first possibility is that massively varying information quantity
    within windows could lead to a small number of such windows having their
    orientation reversed: that is, PC1 becomes PC2 and vice versa. (Or PC2 and PC3
    could switch). This would lead to such windows having unusual properties and
    hence appearing as evidence of an inversion.

    I do agree with the authors that significant outliers would be found at
    inversions. However, even if the PC switching does not occur, or the model could
    handle it, the evidence for selection is weaker.  If the two types of variation
    described above exist, with no selection, I would still expect a ``continuous
    triangle'' of results (as seen left of Fig 2, top left of Fig 6) with extrema
    described by windows with the most information, and points placed at different
    extremum having low recombination rate (because by chance, these will get an
    approximately fixed local tree, corresponding on average to the genome-wide
    population structure).

    Addressing this is likely quite hard, though the authors may be able to think of
    something that separates these effects from selection.
\end{quote}

\begin{point}{}
\ldots variation in the recombination rate \ldots creating spurious structure.
\end{point}

\reply{
    We address this in two ways.
    This point is addressed by comparing results with windows of different types --
    windows of equal length in bp (or in SNPs) have different lengths in cM;
    since these different choices show nearly identical patterns,
    recombination rate variation cannot be driving the results.
}

@@

\reply{

}

\begin{point}{PC switching}
\ldots could lead to a small number of such windows having their
orientation reversed: that is, PC1 becomes PC2 and vice versa. (Or PC2 and PC3
could switch).
\end{point}

\reply{
    This is a natural concern.
    However, the only point at which we compare PCs in a way that could be sensitive to ordering
    is in determining the window size -- in computing the distance between windows
    we use a measure which is invariant under ordering.
    We have made this more clear by moving the note about flipping signs of PCs
    to the appendix on window choice \llname{ll:moved_pc_note}
    and added more explicit notes about this to \llname{ll:another_pc_point} and \revref.
}

@@

\begin{point}{p15}
``We also found nearly identical results when choosing shorter windows of 1,000 SNPs'' - again, show this.
\end{point}

\begin{point}{p15}
 ``or choosing windows of equal length in base pairs rather than SNPs'' - once again.
\end{point}

\begin{point}{}
Using 2 PCs is common practice: only if this is the end of an analysis and the
PCA was done for visualisation. Here you are using it for something so should
keep all the relavant PCs.
\end{point}

\reply{
    This is a good point; the question is which the ``relevant'' PCs are.
    \citet{novembre2008interpreting} showed that under isolation by distance,
    the top two PCs should reflect the two-dimensional nature of the range,
    and higher PCs are generally much less interpretable;
    we used $k=2$ with this in mind.
    We have changed this sentence \revref.
}

@@

%%%%%%%%%%%%%%
\reviewersection{2}

\begin{quote}
    This is an interesting and well written paper. It was a pleasant read. I have three main general
    comments:
\end{quote}

\begin{point}{Related work:} 
The authors provide an introduction of the main concepts, as well as some
intuition of what the method is doing and how, but I found comparison to previous approaches
to be somewhat missing. To some extent, this is due to the fact that the main goal of their
analysis is somewhat vaguely ``finding heterogeneity'', which leads to the applications of
detecting chromosomal inversions and evidence for background selection. It would help to
have a well defined set of hypotheses, test the method’s accuracy  using simulation (see next
comment), and compare to previous efforts in similar domains.
\end{point}

\reply{
    First: we think that ``finding heterogeneity'' is in fact a well-defined goal,
    although it was not that well-defined in the paper;
    we have hopefully improved on this in the Introduction \revref.
    Expanding a bit more:
    We strongly agree that methods that seek to test well-defined hypotheses
    are extremely useful and powerful.
    We also feel that methods for visualization and exploration are also useful --
    a primary example here being PCA.
    If PCA is useful -- and we think that it is --
    then it should be important to also know how much the thing that PCA is summarizing
    varies along the genome,
    in the same way that knowing the mean of some quantity in a population 
    is only of limited usefulness
    without also knowing the corresponding populaion variance.
}

### 2017-06-21 writeup/local_pca_paper.tex
.
reviews; added some recomb stuff

@@

@@

As reviewed above,
it is common to describe variation along the genome of simple statistics such as $F_{ST}$,
and also to use methods such as PCA to visualize large-scale patterns in genome-wide relatedness.
In this paper we 
apply PCA locally along the genome,
thus describing in an unbiased, descriptive way
how patterns of mean relatedness vary systematically along the genome, 
in a way particularly suited to large samples from geographically distributed populations.
Geographic population structure sets the stage by establishing similar patterns of relatedness across much of the genome;
we then aim to describe how this structure is affected by selection and other factors.
Our aim is not to identify outlier loci,
but rather to describe larger-scale variation shared by many parts of the genome.
By doing this with three taxa with diverse population histories,
we can compare relative contributions of different sorts of variation across taxa;
since the method is an visualization method, 
we allow ourselves to be surprised by unexpected signals in the data.

@@

The top few principal components are generally good summaries of population structure; 
we follow common practice and usually only use the first two (referred to as $\pcone$ and $\pctwo$).
The above procedure can be performed on any subset of the data;
for future reference, denote by $\pcone_j$ and $\pctwo_j$
the result after applying to all SNPs in the $j^\text{th}$ window.

Since the definition of eigenvectors does not specify their sign,
when comparing between windows we choose signs to best match each other:
after choosing $\pcone_1$, for instance, 
if $u$ is the first eigenvector obtained from the covariance matrix
for window $j$,
then we next choose $\pcone_j = \pm u$,
where the sign is chosen according to which of 
$\| \pcone_{1} - u \|$ or
$\| \pcone_{1} + u \|$ 
is smaller.

@@

We think of the local effects of population structure as being summarized by \emph{relative} position of the samples
in the space defined by the top principal components.
However, 
we do not compare patterns of relatedness of different genomic regions by directly comparing the PCs,
since rotations or reflections of these imply identical patterns of relatedness.
Instead, we compare the low-dimensional approximations of the local covariance matrices
obtained using the top $k$ PCs,
which is invariant under reflections and rotations and yet contains all other information about the PCs.
(For results shown here, we use $k=2$;
results using larger numbers of PCs were nearly identical.)
Furthermore, to remove the effect of artifacts such as mutation rate variation,
we also rescale each approximate covariance matrix to be of similar size
(precisely, so that the underlying data matrix has trace norm equal to one).

@@

\begin{figure}
    \begin{center}
       \includegraphics[width=0.9\textwidth]{Fig1_allchr_Together_MDS_plot_compact_with_ChrX_inv}
    \end{center}
    \caption{
        Variation in patterns of relatedness for windows across \textit{Drosophila melanogaster} chromosome arms.
        In all plots, each point represents one window along the genome.
         The first column shows the MDS visualization of relationships between windows, 
         and the second and third columns show 
        the two MDS coordinates
         against the midpoint of each window;
         rows correspond to chromosome arms.
         Colors are consistent for plots in each row. 
         Vertical lines show the breakpoints of known polymorphic inversions.   
         Solid black lines are for the inversions we used in \Figure \ref{fig:pca_by_inversion},
         while dotted grey lines are for other known inversions.     
         \label{fig:mds_allchr}
    }
\end{figure}

@@

Unlike the other two species,
the method applied separately on all eight chromosomes of \textit{Medicago truncatula} 
showed similar patterns of gradual change in patterns of relatedness across each chromosome,
with no indications of chromosome-specific patterns.
This consistency suggests that the factor affecting the population structure for each chromosome is the same,
as might be caused by varying strengths of linked selection.
To verify that variation in the effects of population structure is shared across chromosomes,
we applied the method to all chromosomes together.
Results for chromosome 3 are shown in \Figures \ref{fig:mds12_medicago} and \ref{fig:pca_medicago},
and other chromosomes are similar:
% the two types of population structure with higher gene density 
% (colored green and purple in \Figure \ref{fig:mds12_medicago})
% are nearly identical across all chromosomes,
% but population structure associated with low gene density (colored orange)
% shows more chromsome-specific patterns (see Supplemental \Figure \ref{sfig:pca_drosophila_allchr}).
across chromosomes, the high values of the first MDS coordinate coincide with the position of the heterochromatic regions surrounding the centromere,
which often have lower gene density and may therefore be less subject to linked selection.
% Selection on genes in Medicago has been demonstrated by \citet{paape2013selection}. 
To verify that this is a possible explanation,
we counted the number of genes found in each window using gene models in Mt4.0 from \url{jcvi.org} \citep{tang2014improved},
which are shown juxtaposed with
the first MDS coordinate of each window in \Figure \ref{fig:mds_medicago},
and are significantly correlated, as shown in Supplemental \Figure \ref{sfig:mds_gene_count}.
(Values shown are the number of start and end positions of each predicted mRNA transcript,
divided by two, assigned to the nearest window.)
% that lie closer to SNPs  between the last SNP of each window and the first SNP in the next window.
However, other genomic features, such as distance to centromere show roughly the same patterns,
so we cannot rule out alternative hypotheses.
In particular, the recombination rates estimated by \citet{paape2012finescale} appear visually to be similar,
but were not available in a form that can be compared to our MDS values.

@@

Since we summarize patterns of relatedness using relative positions in the principal component maps,
we quantify ``noise'' as the standard error of a sample's position on PC1 in a particular window,
averaged across windows and samples,
and ``signal'' as the standard deviation of the sample's position on PC1 over all windows,
averaged over samples.
(Recall that the signs for PCs are chosen to match each other.)
Then, the mean variance across windows is
\begin{align*}
    \sigma_\text{signal}^2
    = 
    \frac{1}{N} \sum_{j=1}^{N}
        \frac{1}{L}\sum_{i=1}^{L}\left ( \pcone_{ij} -\overline\pcone_{j} \right )^{2} ,
\end{align*}
where $\pcone_{ij}$ is the position of the $i^\text{th}$ individual on $\pcone$ in window $j$,
and $\overline\pcone_j = (1/N) \sum_{j=1}^N \pcone_{ij}$.
We estimate the standard error for each $\pcone_{ij}$ using the block jackknife \citep{efron1982jackknife,busing1999deletem}:
we divide the $j^\text{th}$ window into 10 equal-sized pieces,
and let $\pcone_{ij,k}$ denote the first principal component of this region found after removing the $k^\text{th}$ piece;
then the estimate of the squared standard error is
$\sigma^2_{ij} = \frac{9}{10} \sum_{k=1}^{10} ( \pcone_{ij,k} - \frac{1}{10} \sum_{\ell=1}^{10} \pcone_{ij,\ell} )^2$.
Averaging over samples and windows,
\begin{align*}
    \sigma^2_\text{noise}
    &=
    \frac{1}{N} \sum_{j=1}^{N} \frac{1}{L}\sum_{i=1}^{L} \sigma^2_{ij} .
\end{align*}

@@

\begin{figure}
    \begin{center}
       \includegraphics[width=0.9\textwidth]{FigS_pca_plot_allchr_human}
    \end{center}
    \caption{
        PCA plots for all 22 human autosomes from the POPRES data.
        \label{sfig:pca_human_allchr}
    }
\end{figure}

@@

\begin{figure}
    \begin{center}
        \includegraphics{MDS_allchr_Together_plot_samples_noinv}
    \end{center}
    \caption{
        Variation in structure for windows of 1,000 SNPs 
        across \textit{Drosophila melanogaster} chromosome arms: without inversions.
         As in \Figure \ref{fig:mds_allchr}, but after omitting for each chromosome arm individuals carrying the less frequent orientation
         of any inversions on that chromosome arm.
         The values differ from those in \ref{fig:drosophila_recomb_rate}
         in the window size used and that some MDS values were inverted
         (but relative orientation is meaningless as chromosome arms were run separately,
         unlike for \textit{Medicago}).
        In all plots, each point represents one window along the genome.
         The first column shows the MDS visualization of relationships between windows, 
         and the second and third columns show the midpoint of each window against the two MDS coordinates; 
         rows correspond to chromosome arms.
         Colors are consistent for plots in each row. 
         Vertical lines show the breakpoints of known polymorphic inversions.   
         \label{sfig:mds_allchr_noinversions}
    }
\end{figure}

@@

\begin{figure}
    \begin{center}
       \includegraphics[width=0.85\textwidth]{FigS_Together_MDS_plot_allchr}
    \end{center}
    \caption{
        MDS visualizations of the effects of population structure for all 8 chromosomes of the \textit{Medicago truncatula} data.
        \label{sfig:mds_medicago_allchr}
    }
\end{figure}

>>>>

% some figs with many points make the pdfs render slow
\newif\ifusepngs
\usepngstrue

@@

To-do: cite \citet{duforetfrebourg2015detecting}.

@@

It is common to describe variation along the genome of simple statistics such as $F_{ST}$
and to interpret the results in terms of the action of selection 
\citep{speciation_islands,differentiation_islands,flycatchers}.
However, it has been pointed out that a given pattern (e.g., valleys of $F_{ST}$)
can be caused by more than one biological process \citep{hahn_fst,flycatchers2},
which in retrospect is unsuprising given that we are using a single statistic
to describe a complex process.
It is also common to use methods such as PCA to visualize large-scale patterns in 
mean genome-wide relatedness.
In this paper we show if and how patterns of mean relatedness vary systematically along the genome, 
in a way particularly suited to large samples from geographically distributed populations.
Geographic population structure sets the stage by establishing ``background'' patterns of relatedness;
our method then describes how this structure is affected by selection and other factors.
Our aim is not to identify outlier loci,
but rather to describe larger-scale variation shared by many parts of the genome;
correlation of this variation with known genomic features
can then be used to uncover its source. \revpoint{2}{1}

@@

The top 2--5 principal components are generally good summaries of population structure; 
for ease of visualization we usually only use the first two (referred to as $\pcone$ and $\pctwo$),
and check that results hold using more. \revpoint{1}{7}
The above procedure can be performed on any subset of the data;
for future reference, denote by $\pcone_j$ and $\pctwo_j$
the result after applying to all SNPs in the $j^\text{th}$ window.
(Note, however, that our measure of dissimilarity between windows
does not depend on PC ordering.) \linelabel{ll:another_pc_point}

@@

We think of the local effects of population structure as being summarized by \emph{relative} position of the samples
in the space defined by the top principal components.
However, 
we do not compare patterns of relatedness of different genomic regions by directly comparing the PCs,
since rotations or reflections of these imply identical patterns of relatedness.
Instead, we compare the low-dimensional approximations of the local covariance matrices
obtained using the top $k$ PCs,
which is invariant under ordering of the PCs \revpoint{1}{3},
reflections, and rotations and yet contains all other information about the PCs.
(For results shown here, we use $k=2$;
results using larger numbers of PCs were nearly identical.)
Furthermore, to remove the effect of artifacts such as mutation rate variation,
we also rescale each approximate covariance matrix to be of similar size
(precisely, so that the underlying data matrix has trace norm equal to one).

@@

\begin{figure}
    \begin{center}
        \ifusepngs
           \includegraphics[width=0.9\textwidth]{Fig1_allchr_Together_MDS_plot_compact_with_ChrX_inv.png}
       \else
           \includegraphics[width=0.9\textwidth]{Fig1_allchr_Together_MDS_plot_compact_with_ChrX_inv}
       \fi
    \end{center}
    \caption{
        Variation in patterns of relatedness for windows across \textit{Drosophila melanogaster} chromosome arms.
        In all plots, each point represents one window along the genome.
         The first column shows the MDS visualization of relationships between windows, 
         and the second and third columns show 
        the two MDS coordinates
         against the midpoint of each window;
         rows correspond to chromosome arms.
         Colors are consistent for plots in each row. 
         Vertical lines show the breakpoints of known polymorphic inversions.   
         Solid black lines are for the inversions we used in \Figure \ref{fig:pca_by_inversion},
         while dotted grey lines are for other known inversions.     
         \label{fig:mds_allchr}
    }
\end{figure}

@@

Unlike the other two species,
the method applied separately on all eight chromosomes of \textit{Medicago truncatula} 
showed similar patterns of gradual change in patterns of relatedness across each chromosome,
with no indications of chromosome-specific patterns.
This consistency suggests that the factor affecting the population structure for each chromosome is the same,
as might be caused by varying strengths of linked selection.
To verify that variation in the effects of population structure is shared across chromosomes,
we applied the method to all chromosomes together.
Results for chromosome 3 are shown in \Figures \ref{fig:mds12_medicago} and \ref{fig:pca_medicago},
and other chromosomes are similar:
% the two types of population structure with higher gene density 
% (colored green and purple in \Figure \ref{fig:mds12_medicago})
% are nearly identical across all chromosomes,
% but population structure associated with low gene density (colored orange)
% shows more chromsome-specific patterns (see Supplemental \Figure \ref{sfig:pca_drosophila_allchr}).
across chromosomes, the high values of the first MDS coordinate coincide with the position of the heterochromatic regions surrounding the centromere,
which often have lower gene density and may therefore be less subject to linked selection.
% Selection on genes in Medicago has been demonstrated by \citet{paape2013selection}. 
To verify that this is a possible explanation,
we counted the number of genes found in each window using gene models in Mt4.0 from \url{jcvi.org} \citep{tang2014improved},
which are shown juxtaposed with
the first MDS coordinate of each window in \Figure \ref{fig:mds_medicago},
and are significantly correlated, as shown in Supplemental \Figure \ref{sfig:mds_gene_count}.
(Values shown are the number of start and end positions of each predicted mRNA transcript,
divided by two, assigned to the nearest window.)
% that lie closer to SNPs  between the last SNP of each window and the first SNP in the next window.
However, other genomic features, such as distance to centromere show roughly the same patterns,
so we cannot rule out alternative hypotheses.
In particular, the recombination rates estimated by \citet{paape2012finescale} appear visually to be similar,
but were not available in a form that can be reliably compared to our MDS values. \revpoint{2}{8}

@@

Since we summarize patterns of relatedness using relative positions in the principal component maps,
we quantify ``noise'' as the standard error of a sample's position on PC1 in a particular window,
averaged across windows and samples,
and ``signal'' as the standard deviation of the sample's position on PC1 over all windows,
averaged over samples.
The definition of eigenvectors does not specify their sign,
and so when comparing between windows we choose signs to best match each other:
after choosing $\pcone_1$, for instance, 
if $u$ is the first eigenvector obtained from the covariance matrix
for window $j$,
then we next choose $\pcone_j = \pm u$,
where the sign is chosen according to which of 
$\| \pcone_{1} - u \|$ or
$\| \pcone_{1} + u \|$ 
is smaller.
\linelabel{ll:moved_pc_note}

After doing this, the mean variance across windows is
\begin{align*}
    \sigma_\text{signal}^2
    = 
    \frac{1}{N} \sum_{j=1}^{N}
        \frac{1}{L}\sum_{i=1}^{L}\left ( \pcone_{ij} -\overline\pcone_{j} \right )^{2} ,
\end{align*}
where $\pcone_{ij}$ is the position of the $i^\text{th}$ individual on $\pcone$ in window $j$,
and $\overline\pcone_j = (1/N) \sum_{j=1}^N \pcone_{ij}$.
We estimate the standard error for each $\pcone_{ij}$ using the block jackknife \citep{efron1982jackknife,busing1999deletem}:
we divide the $j^\text{th}$ window into 10 equal-sized pieces,
and let $\pcone_{ij,k}$ denote the first principal component of this region found after removing the $k^\text{th}$ piece;
then the estimate of the squared standard error is
$\sigma^2_{ij} = \frac{9}{10} \sum_{k=1}^{10} ( \pcone_{ij,k} - \frac{1}{10} \sum_{\ell=1}^{10} \pcone_{ij,\ell} )^2$.
Averaging over samples and windows,
\begin{align*}
    \sigma^2_\text{noise}
    &=
    \frac{1}{N} \sum_{j=1}^{N} \frac{1}{L}\sum_{i=1}^{L} \sigma^2_{ij} .
\end{align*}

@@

\begin{figure}
    \begin{center}
        \ifusepngs
           \includegraphics[width=0.9\textwidth]{FigS_pca_plot_allchr_human.png}
       \else
           \includegraphics[width=0.9\textwidth]{FigS_pca_plot_allchr_human}
       \fi
    \end{center}
    \caption{
        PCA plots for all 22 human autosomes from the POPRES data.
        \label{sfig:pca_human_allchr}
    }
\end{figure}

@@

\begin{figure}
    \begin{center}
        \ifusepngs
            \includegraphics{MDS_allchr_Together_plot_samples_noinv.png}
        \else
            \includegraphics{MDS_allchr_Together_plot_samples_noinv}
        \fi
    \end{center}
    \caption{
        Variation in structure for windows of 1,000 SNPs 
        across \textit{Drosophila melanogaster} chromosome arms: without inversions.
         As in \Figure \ref{fig:mds_allchr}, but after omitting for each chromosome arm individuals carrying the less frequent orientation
         of any inversions on that chromosome arm.
         The values differ from those in \ref{fig:drosophila_recomb_rate}
         in the window size used and that some MDS values were inverted
         (but relative orientation is meaningless as chromosome arms were run separately,
         unlike for \textit{Medicago}).
        In all plots, each point represents one window along the genome.
         The first column shows the MDS visualization of relationships between windows, 
         and the second and third columns show the midpoint of each window against the two MDS coordinates; 
         rows correspond to chromosome arms.
         Colors are consistent for plots in each row. 
         Vertical lines show the breakpoints of known polymorphic inversions.   
         \label{sfig:mds_allchr_noinversions}
    }
\end{figure}

@@

\begin{figure}
    \begin{center}
        \ifusepngs
           \includegraphics[width=0.85\textwidth]{FigS_Together_MDS_plot_allchr.png}
       \else
           \includegraphics[width=0.85\textwidth]{FigS_Together_MDS_plot_allchr}
       \fi
    \end{center}
    \caption{
        MDS visualizations of the effects of population structure for all 8 chromosomes of the \textit{Medicago truncatula} data.
        \label{sfig:mds_medicago_allchr}
    }
\end{figure}
